---
title: "Article Plotting Notebook"
output: html_notebook
---

This Notebook ooutlines the code required to create the plot of different samples
of nutrients tested in water around the world. This plot will be used in an article
that will hopefully be published in the Canadian Journal of Chemical Engineering. 

The following code is used to install the required libraries. 
```{r, Initial}
library(tidyverse)
library(dataRetrieval)
library(ggmap)
library(maps)
library(mapdata)
library(mapsapi)
```

This next chunk of code is uses the readWQPdata function from the datRetriveal package
to download data from the Water Quality Portal. The Water Quality Portal is a collection
of three datasets of water quality sampling across the globe. 

The parameters are passed to the function to define the scope for the data to collection,
for this instance it limits the window for when the sample was first collected and
limits the database to solely the concentration of total Phosphorus.

```{r, dataget}
wqp.data <- readWQPdata('startDateLo' = "2016-01-01", 
                        'startDateHi' = "2016-12-31", 
                        characteristicName=c("Phosphorus")) %>%
        filter(ResultSampleFractionText=="Total")
```


